---
title: "EDA"
author: "Tilina Alzaben"
date: "2023-09-27"
output: html_document
---

**Introduction** 

In the late 19th and early 20th centuries, the United States saw a massive expansion of its railway system, connecting towns and cities for both people and goods. However, things changed with the rise of cars and the Interstate Highway System, which encouraged long-distance road travel. As a result, railways became less popular, leading to the abandonment of many rail lines. Interestingly, some of these disused railway lines found new life as "rail trails" – paths for walking and biking that are easy to access and great for outdoor activities.

Our study aims to investigate whether living near a rail trail makes homes more appealing and whether people are willing to pay extra for such homes. To do this, we're using data from 104 homes sold in Northampton, Massachusetts, in 2007. This research can help Acme Homes, LLC, a company that builds and sells houses, find promising locations for future housing projects. We'll analyze the data to see if there's a connection between proximity to a rail trail and home values, helping Acme Homes make informed decisions about where to build homes.

**Exploratory Data Analysis**

This dataset includes a wide range of information about residential properties, including details such as their values, sizes, bedroom counts, proximity to rail trails, bike and walkability scores, and more. Our goal in analyzing this dataset is to gain a deeper understanding of its underlying characteristics, providing a foundation for more comprehensive analysis.

The dataset not only offers information on home values but also tracks changes in these values over various years. Additionally, it provides location-specific data, including latitude, longitude, and ZIP codes. This geographical data is particularly valuable for identifying potential spatial patterns within the dataset.

One key aspect we investigate is the distance of each home to the nearest rail trail, expressed both in feet and as a categorical variable ("Closer" or "Farther Away"). This approach allows us to explore how living in proximity to a rail trail may impact property values. During our analysis, we've noticed a few outliers in the data, notably a $669,925 home closer to a rail trail and several homes with values of $798,624, $516,520, $513,647, and $495,836, which are farther away from rail trails. These outliers could provide valuable insights into the relationship between distance to rail trails and property values. 

```{r}
ggplot(houses, aes(x = distgroup, y = adj2007)) +
  geom_boxplot(fill = "steelblue", color = "darkblue") +
  labs(title = "House Values by Proximity to Rail Trail",
       x = "Distance to Rail Trail",
       y = "House Value (thousands)")
```

Moreover, the dataset contains essential information regarding property attributes, such as acres or square footage, bedroom counts, bathroom numbers, and the availability of garage parking. This wealth of data enables us to delve into the factors that exert influence on property values, shedding light on the key determinants.

Additionally, we have bike and walking scores within the dataset which serve as indicators of the neighborhood's bike-friendliness and walkability. Both scatterplots below demonstrate that bikescore and walkscore exhibit remarkably similar patterns, suggesting a strong correlation. It's worth noting that these scores are third-party evaluations and may take into account factors like proximity to rail trails. Consequently, there is the possibility that these scores are not entirely independent, a relationship we will explore in more depth in subsequent sections of our analysis.

```{r}
library(patchwork)

bikescore <- ggplot(houses, aes(x = bikescore, y = distance)) +
  geom_point(color = "darkblue") +
  labs(title = "Bike Score and Distance to Rail Trail",
       y = "Distance to Rail Trail (in feet)",
       x = "Bike Score")

walkscore <- ggplot(houses, aes(x = walkscore, y = distance)) +
  geom_point(color = "steelblue") +
  labs(title = "Walk Score and Distance to Rail Trail",
       y = "Distance to Rail Trail (in feet)",
       x = "Walk Score")

bikescore + walkscore
```

In the course of our analysis, we uncovered some noteworthy findings. For example, in the year 2007, the average house value was approximately $327,600. This figure provides valuable insight into the typical property values for that specific year, offering a crucial reference point. Furthermore, our examination revealed that the average distance from homes to rail trails is roughly 1.11 feet. This metric provides a clear picture of how proximate these properties are to these recreational pathways, highlighting their accessibility.

```{r, echo = FALSE, message = FALSE}

houses %>%
  ggplot(aes(x = adj2007)) +
  geom_histogram(fill = "steelblue", col = "darkblue") +
  geom_vline(aes(xintercept = mean(adj2007)), col = "red") +
  theme_bw() +
  labs(
    title = "Distribution of House Values in 2007",
    subtitle = "(the average house value cost $327,600)",
    x = "House Value in 2007 (in thousands)",
    y = "Count"
  )

houses %>%
  ggplot(aes(x = distance)) +
  geom_histogram(fill = "steelblue", col = "darkblue") +
  geom_vline(aes(xintercept = mean(distance)), col = "red") +
  theme_bw() +
  labs(
    title = "Distribution of Distance to Rail Trail",
    subtitle = "(the average distance is 1.11 ft)",
    x = "Distance (in feet)",
    y = "Count"
  )
```
Additionally, our analysis unveiled an intriguing correlation: the distance to a rail trail exhibited a negative correlation with home value. This suggests that there is a potential relationship between distance from a rail trail and home value. Specifically, it implies that homes situated farther away from rail trails may tend to be more affordable. This correlation warrants further exploration to better understand its implications.

```{r}
houses %>%
  ggplot(aes(x = distance, y = adj2007)) +
  geom_point(col = "steelblue") +
  theme_bw() +
  labs(
    title = "Scatterplot of House Values in 2007 vs. Distance",
    x = "Distance (in feet)",
    y = "House Values in 2007 (in thousands)"
  )
```
Furthermore, it's worth noting that the majority of homes within the dataset boast three bedrooms and either zero or one garage space, indicating prevalent features among these residential properties. This observation provides valuable insight into the common characteristics of the homes under consideration.

Importantly, one key highlight of our analysis is the absence of missing values in the dataset. This means that we have complete and comprehensive data for all variables, ensuring the reliability and integrity of our dataset throughout our analytical process.

```{r}
bedrooms <- houses %>%
  ggplot(aes(x = bedrooms)) +
  geom_bar(fill = "steelblue", color = "darkblue") +
  theme_bw() +
  labs(
    title = "Number of Bedrooms",
    x = "Bedrooms",
    y = "Count"
  )

garage <- houses %>%
  ggplot(aes(x = garage_spaces)) +
  geom_bar(fill = "steelblue", color = "darkblue") +
  theme_bw() +
  labs(
    title = "Number of Garage Space",
    x = "Garages",
    y = "Count"
  )

bedrooms + garage
```

In our data preprocessing phase, we made some essential adjustments to improve the dataset's quality and usability. We created a new variable, "no_baths," by combining bathroom information in a way that simplifies our analysis. We also treated ZIP codes as categorical, recognizing that they are not numbers with an inherent order. Furthermore, we treated the "no_bedrooms" variable as categorical to make it easier to understand. Lastly, we improved variable names to make them more intuitive. These preprocessing steps help make the dataset more suitable for in-depth analysis while preserving the data's accuracy and meaning.

```{r}
houses %>%
  ggplot(aes(x = bedrooms, y = adj2007)) +
  geom_point(col = houses$bedgroup, alpha = 0.5) + 
  theme_bw()+
  labs(
    title = "Scatterplot of House Values in 2007 vs. Number of Bedrooms",
    x = "Bedrooms",
    y = "House Values in 2007 (in thousands)"
  )
```




