---
title: "Project 1: Rail Trails"
author: "Gabriel Krotkov"
date: "2023-09-25"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r data loading, echo = FALSE, message = FALSE, warning = FALSE}
library(tidyverse)
library(gridExtra)

houses <- suppressMessages(read_csv("JSE13-070R2.csv"))

# for simplicity, combine full and half baths
houses$no_baths <- houses$no_full_baths + (0.5 * houses$no_half_baths)

houses$zip <- factor(houses$zip)

houses <- houses[, c("acre", "adj2007", "bedrooms", "bikescore", 
                  "walkscore", "distance", "garage_spaces", "no_rooms", 
                  "no_baths", "squarefeet", "zip")]

save(houses, file = "data.rda")
```

```{r}
fit1 <- lm(log(adj2007) ~ log(distance) + bedrooms + garage_spaces + no_baths + zip + 
               squarefeet + no_rooms + acre, 
           data = houses)

summary(fit1)

plot(fit1)
```

Reasons for including/not including variables: 
- Walkscore/bikescore, do NOT include, we think they overlap with `distance` in that they incorporate the nearby presence of the Rail Trail
- 

# Introduction 

We will write an introduction here 

# Exploratory Data Analysis

We will integrate the EDA from our three different projects here

# Methods 



# Results 

We will show plots and model summaries here

# Discussions 

We will discuss our overall results and come to a conclusion here.